---
title: "Correlation vs. Cosine Similarity"
subtitle: "50 ML Projects to Understand LLMs — Miscellaneous<br>R translation of Mike X Cohen's Substack post"
author: "Dan Killian, with Posit AI<br>an R translation of Mike X Cohen's original Python notebook"
date: today
format:
  html:
    toc: true
    code-fold: false
    theme: cosmo
execute:
  warning: false
  message: false
---

*Using the code without reading the post may lead to confusion or errors.*

---

```{r global_options, include=F, warning=F, message=F, echo=F, error=F}
library(here)
library(knitr)
source("../../../prep (May 2025).r")
```

---

# Setup

We need `ggplot2` for visualization and `patchwork` to assemble multi-panel figures. Both are loaded via the prep script — confirm they're available before proceeding.

```{r}
#| label: setup
#| eval: false

# Load the two libraries you'll use throughout this exercise.
# Hint: one for plotting, one for combining plots side-by-side.
library(___)
library(___)
```

---

# Illustration of Data Simulation

We'll start by building intuition for correlation as a geometric concept. The idea: given a desired population correlation $$\rho$$, we can generate a pair of random variables $$(x, y)$$ where $$x$$ is standard normal and $$y$$ is constructed to have exactly that linear relationship with $$x$$, plus independent noise.

The formula for generating $$y$$ from $$x$$ with correlation $$\rho$$ is:

$$y = x \cdot \rho + \varepsilon \cdot \sqrt{1 - \rho^2}$$

where $$\varepsilon \sim \mathcal{N}(0,1)$$ is independent noise. The $$\sqrt{1 - \rho^2}$$ term scales the noise so that the total variance of $$y$$ stays at 1.

::: {.callout-tip title="Exercise prompts"}
-   What does it mean for two variables to have a population correlation of $$\rho$$?
-   In the formula above, what happens to $$y$$ when $$\rho = 1$$? When $$\rho = 0$$?
-   We want three scatter plots side-by-side, one each for $$\rho \in \{-0.7,\ 0,\ 0.7\}$$. How would you store and iterate over those values in R?
-   For each subplot, how do you compute the *sample* correlation from the data to display in the title?
:::

```{r}
#| label: fig-sim-illustration
#| eval: false
#| fig-width: 10
#| fig-height: 3.5

set.seed(___)  # choose any seed

pop_cors <- c(___, ___, ___)  # three target correlations
N        <- 90                # sample size

plots <- lapply(pop_cors, function(rho) {

  x <- rnorm(___)
  y <- ___  # use the formula above

  r <- cor(___, ___)  # sample correlation

  df <- data.frame(x = x, y = y)

  ggplot(df, aes(x, y)) +
    geom_point() +
    labs(
      x     = 'Data "x"',
      y     = 'Data "y"',
      title = sprintf("Correlation = %.2f", ___)
    )
})

wrap_plots(plots, nrow = 1)
```

---

# Correlations Can Mislead

Pearson's $$r$$ only captures *linear* association. Three classic shapes produce $$r \approx 0$$ even though the relationship between $$x$$ and $$y$$ is perfectly deterministic:

| Shape | x | y |
|---|---|---|
| Circle | $$\cos(\theta)$$ | $$\sin(\theta)$$ |
| Parabola | $$x$$ | $$x^2$$ |
| V-shape | $$x$$ repeated twice | $$x$$ then $$-x$$ |

::: {.callout-tip title="Exercise prompts"}
-   For the circle: what range of angles $$\theta$$ traces a full circle? Use `seq()` — be careful not to repeat the endpoint.
-   For the parabola: generate evenly spaced $$x$$ values over $$[-2, 2]$$. What is $$y$$?
-   For the V-shape: you need $$x$$ to repeat itself and $$y$$ to be $$x$$ stacked with $$-x$$. What R function concatenates vectors?
-   After generating each shape, compute $$r$$ and display it in the plot title.
:::

```{r}
#| label: fig-misleading-cors
#| eval: false
#| fig-width: 10
#| fig-height: 3.5

N <- 90

make_shape <- function(shape_id) {

  if (shape_id == 1) {
    # Circle: theta from 0 to 2*pi (don't include 2*pi itself)
    theta <- seq(0, ___, length.out = N)
    x <- cos(___)
    y <- sin(___)

  } else if (shape_id == 2) {
    # Parabola
    x <- seq(-2, 2, length.out = N)
    y <- ___

  } else {
    # V-shape: x goes from -2 to 2 (N/2 points), repeated; y = x then -x
    half <- seq(-2, 2, length.out = N %/% 2)
    x <- c(___, ___)
    y <- c(___, ___)
  }

  r <- cor(x, y)
  data.frame(x = x, y = y, r = r, shape = shape_id)
}

shapes <- lapply(1:3, make_shape)

plots <- lapply(shapes, function(df) {
  ggplot(df, aes(x, y)) +
    geom_point() +
    labs(title = sprintf("r = %.2f", df$r[1])) +
    theme(axis.text = element_blank(), axis.ticks = element_blank(),
          axis.title = element_blank())
})

wrap_plots(plots, nrow = 1)
```

---

# Systematic Comparison of Correlation and Cosine Similarity

Now for the main event. Both metrics measure how "aligned" two vectors are — but they differ in one important way.

**Pearson correlation** centers each variable before computing alignment:

$$r = \frac{\sum (x_i - \bar{x})(y_i - \bar{y})}{\sqrt{\sum (x_i - \bar{x})^2 \cdot \sum (y_i - \bar{y})^2}}$$

**Cosine similarity** uses the raw vectors:

$$S_C = \frac{\sum x_i y_i}{\sqrt{\sum x_i^2 \cdot \sum y_i^2}}$$

The only difference is whether you subtract the mean before computing. This makes cosine similarity sensitive to the *location* of the data, not just its shape. We'll demonstrate this by adding a large mean offset to $$x$$ before computing both metrics.

::: {.callout-tip title="Exercise prompts"}
-   Loop over 100 evenly spaced population correlations from $$-1$$ to $$1$$. For each:
    1.  Generate $$(x, y)$$ with $$N = 500$$ using the same formula as before.
    2.  Shift $$x$$ by subtracting 10 from every value. This is the mean offset.
    3.  Compute $$r$$ using the Pearson formula (subtract means first). You can check your result with `cor()`.
    4.  Compute $$S_C$$ using the cosine formula (no mean subtraction). You can check with `lsa::cosine()` if you have the package.
-   Store both metrics for each iteration, then visualize:
    -   Left plot: population $$\rho$$ on the x-axis, both $$r$$ and $$S_C$$ on the y-axis.
    -   Right plot: $$r$$ on the x-axis, $$S_C$$ on the y-axis. What pattern do you expect?
:::

```{r}
#| label: fig-systematic-comparison
#| eval: false
#| fig-width: 10
#| fig-height: 4

set.seed(___)

pop_rs <- seq(-1, 1, length.out = ___)  # 100 evenly spaced values
N      <- 500

results <- lapply(pop_rs, function(rho) {

  x <- rnorm(N)
  y <- ___            # generate y from rho

  x <- x - ___       # apply mean offset

  # Pearson correlation (from formula, not cor())
  x_c   <- x - mean(x)
  y_c   <- y - mean(y)
  x_ss  <- sum(x_c^2)
  y_ss  <- ___
  r     <- sum(___) / sqrt(___ * ___)

  # Cosine similarity (from formula)
  x_ss2 <- sum(x^2)
  y_ss2 <- ___
  cos_s <- sum(___) / sqrt(___ * ___)

  data.frame(pop_r = rho, pearson = r, cosine = cos_s)
})

res <- do.call(rbind, results)

# Left plot: both metrics vs. population rho
p1 <- ggplot(res, aes(x = pop_r)) +
  geom_point(aes(y = ___, shape = "Correlation")) +
  geom_point(aes(y = ___, shape = "Cosine sim.")) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray60") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray60") +
  scale_shape_manual(values = c("Correlation" = 15, "Cosine sim." = 16)) +
  labs(
    x     = "Population correlation",
    y     = expression("Measured " * italic(r) * " or " * italic(S[C])),
    title = "Correlation and cosine sim.",
    shape = NULL
  )

# Right plot: Pearson r vs. cosine similarity
r_between <- cor(res$pearson, res$cosine)

p2 <- ggplot(res, aes(x = ___, y = ___)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray60") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray60") +
  coord_cartesian(xlim = c(-1.1, 1.1), ylim = c(-1.1, 1.1)) +
  labs(
    x     = "Correlation",
    y     = "Cosine similarity",
    title = sprintf("r = %.2f", r_between)
  )

___ | ___   # combine with patchwork
```
