---
title: "Book Lengths in Characters, Words, and Tokens"
subtitle: "50 ML Projects to Understand LLMs — Chapter 2, Project 2"
author: "Dan Killian, with Posit AI<br>an R translation of Mike X Cohen's original Python notebook"
date: today
format:
  html:
    toc: true
    code-fold: false
    theme: cosmo
execute:
  warning: false
  message: false
---

*Using the code without reading the book may lead to confusion or errors.*

---

```{r global_options, include=F, warning=F, message=F, echo=F, error=F}
library(here)
library(knitr)
source("../../../prep (May 2025).r")
```

---

# How to Import a Book from gutenberg.org

Project Gutenberg offers thousands of public-domain texts in plain text format. Each book is accessible via a URL with a consistent structure — a base URL plus a numeric code unique to each work.

```{r}
#| label: gutenberg-demo
url  <- "https://www.gutenberg.org/cache/epub/84/pg84.txt"
text <- readLines(url, encoding = "UTF-8") |> paste(collapse = "\n")
cat(substr(text, 1, 500))
```

---

# Setup: Books and Tokenizer

We select 10 classic books and load the GPT-2 tokenizer as a third comparison alongside character- and word-based tokenization.

```{r}
#| label: books-and-tokenizer
baseurl <- "https://www.gutenberg.org/cache/epub/"

books <- data.frame(
  code  = c("84",    "64317",       "11",              "1513",        "76",
            "219",          "2591",         "2148",           "36",              "829"),
  title = c("Frankenstein", "GreatGatsby", "AliceWonderland", "RomeoJuliet", "HuckFinn",
            "HeartDarkness", "GrimmsTales", "EdgarAllenPoe", "WarOfTheWorlds", "GulliversTravels")
)

# Load GPT-2 tokenizer (downloads vocab on first run)
gpt2 <- tok::tokenizer$from_pretrained("gpt2")
```

---

# Part 1: Total Characters, Words, and Tokens

For each book, we count total characters (including spaces, punctuation, and newlines), whitespace-delimited words, and GPT-2 tokens.

```{r}
#| label: part1-counts
#| cache: true
results <- vector("list", nrow(books))

for (i in seq_len(nrow(books))) {
  fullurl <- paste0(baseurl, books$code[i], "/pg", books$code[i], ".txt")
  text    <- readLines(fullurl, encoding = "UTF-8") |> paste(collapse = "\n")

  results[[i]] <- data.frame(
    title   = books$title[i],
    n_chars = nchar(text),
    n_words = length(strsplit(trimws(text), "\\s+")[[1]]),
    n_token = length(gpt2$encode(text)$ids)
  )
}

counts <- bind_rows(results)
```

```{r}
#| label: tbl-part1
counts %>%
  flextable() %>%
  flextable::set_caption("Total character, word, and token counts for 10 Project Gutenberg books")
```

Token counts track word counts fairly closely. GPT-2 subword tokenization produces more tokens than words, since punctuation and uncommon words often split into multiple pieces.

---

# Part 2: Unique and Total Counts

Beyond totals, how many *distinct* characters, words, and tokens appear in each book? This is a rough measure of vocabulary richness.

```{r}
#| label: part2-unique
#| cache: true
full_results <- vector("list", nrow(books))

for (i in seq_len(nrow(books))) {
  fullurl <- paste0(baseurl, books$code[i], "/pg", books$code[i], ".txt")
  text    <- readLines(fullurl, encoding = "UTF-8") |> paste(collapse = "\n")

  chars  <- strsplit(text, "")[[1]]
  words  <- strsplit(trimws(text), "\\s+")[[1]]
  tokens <- gpt2$encode(text)$ids

  full_results[[i]] <- data.frame(
    title        = books$title[i],
    total_chars  = length(chars),
    unique_chars = length(unique(chars)),
    total_words  = length(words),
    unique_words = length(unique(words)),
    total_token  = length(tokens),
    unique_token = length(unique(tokens))
  )
}

full_counts <- bind_rows(full_results)
```

```{r}
#| label: tbl-part2
full_counts %>%
  flextable() %>%
  flextable::set_caption("Total and unique counts by tokenization scheme for 10 books")
```

```{r}
#| label: fig-part2
#| fig-cap: "Total (A) and unique (B) counts across tokenization schemes for 10 Project Gutenberg books. Points are individual books; crossbar marks the mean."
#| fig-width: 10
#| fig-height: 3.5

# Reshape to long format for plotting
long_counts <- full_counts |>
  pivot_longer(
    cols      = -title,
    names_to  = c("measure", "scheme"),
    names_sep = "_",
    values_to = "count"
  ) |>
  mutate(
    scheme  = factor(tools::toTitleCase(scheme), levels = c("Chars", "Words", "Token")),
    measure = if_else(measure == "total", "A) Total", "B) Unique")
  )

ggplot(long_counts, aes(x = scheme, y = count)) +
  geom_jitter(width = 0.08, size = 2.5,
              color = "dodgerblue2", alpha = 0.8) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.6,
               linewidth = 0.8, color = "blue") +
  scale_y_continuous(labels = scales::label_scientific()) +
  facet_wrap(~ measure, scales = "free_y") +
  labs(x = NULL, y = "Count") +
  faceted +
  theme(strip.background = element_blank(),
        strip.text        = element_text(face = "bold"))
```

Unique character counts are tightly clustered — most books draw from roughly the same set of ASCII characters. Unique word and token counts spread out more, reflecting differences in author vocabulary and subject matter.

---

# Part 3: Unique by Total Counts

Do longer books use a wider vocabulary? We examine the relationship between total and unique counts across the three tokenization schemes.

```{r}
#| label: fig-part3
#| fig-cap: "Relationship between unique and total counts for each tokenization scheme across 10 books."
#| fig-width: 12
#| fig-height: 3.5

# Reshape: one row per book per scheme, with columns for total and unique
scatter_data <- full_counts |>
  select(title,
         chars_total  = total_chars,  chars_unique  = unique_chars,
         words_total  = total_words,  words_unique  = unique_words,
         token_total  = total_token,  token_unique  = unique_token) |>
  pivot_longer(
    cols      = -title,
    names_to  = c("scheme", "measure"),
    names_sep = "_"
  ) |>
  pivot_wider(names_from = measure, values_from = value) |>
  mutate(scheme = factor(tools::toTitleCase(scheme),
                         levels = c("Chars", "Words", "Token")))

# Per-scheme correlation labels
cor_labels <- scatter_data |>
  group_by(scheme) |>
  summarise(r = cor(unique, total), .groups = "drop") |>
  mutate(label = sprintf("r = %.2f", r))

ggplot(scatter_data, aes(x = unique, y = total)) +
  geom_point(size = 3, color = "dodgerblue2") +
  geom_text(data = cor_labels, aes(label = label),
            x = -Inf, y = Inf, hjust = -0.2, vjust = 1.5,
            fontface = "bold", inherit.aes = FALSE) +
  scale_x_continuous(labels = scales::label_scientific()) +
  scale_y_continuous(labels = scales::label_scientific()) +
  facet_wrap(~ scheme, scales = "free") +
  labs(x = "Unique", y = "Total") +
  faceted +
  theme(strip.background = element_blank(),
        strip.text        = element_text(face = "bold"))
```

Characters show essentially no relationship between unique and total counts — the character set saturates quickly regardless of book length. Word and token unique counts track total counts more closely, suggesting longer books do tend to draw on larger vocabularies.

One book appears as an outlier in unique character count. Let's identify it.

```{r}
#| label: outlier
full_counts |>
  slice_max(unique_chars, n = 1) |>
  select(title, unique_chars) %>%
  flextable()
```

The Edgar Allen Poe collection contains an unusually large number of distinct characters. We can inspect the full character set to understand why.

```{r}
#| label: outlier-chars
text_poe <- readLines(
  "https://www.gutenberg.org/cache/epub/2148/pg2148.txt",
  encoding = "UTF-8"
) |> paste(collapse = "\n")

sort(unique(strsplit(text_poe, "")[[1]]))
```

The Poe collection includes accented and non-Latin characters — likely from French or other language quotations embedded in the text — which inflates the unique character count relative to the other books.
