---
title: "Book Lengths in Characters, Words, and Tokens"
subtitle: "50 ML Projects to Understand LLMs — Chapter 2, Project 2"
author: "Dan Killian, with Posit AI<br>an R translation of Mike X Cohen's original Python notebook"
date: today
format:
  html:
    toc: true
    code-fold: false
    theme: cosmo
    fig-height: 3
    fig-width: 5
execute:
  warning: false
  message: false
---

*Using the code without reading the book may lead to confusion or errors.*

---

```{r global_options, include=F, warning=F, message=F, echo=F, error=F, fig.height=3, fig.width=5}
library(here)
library(knitr)
source("../../../prep (May 2025).r")
theme_set(base_quarto)
```

---

# How to Import a Book from gutenberg.org

Project Gutenberg offers thousands of public-domain texts in plain text format. Each book is accessible via a URL with a consistent structure — a base URL plus a numeric code unique to each work.

```{r}
#| label: gutenberg-demo
url  <- "https://www.gutenberg.org/cache/epub/84/pg84.txt"
text <- readLines(url, encoding = "UTF-8") |> paste(collapse = "\n")
cat(substr(text, 1, 500))
```

---

# Setup: Books and Tokenizer

We select 10 classic books and load the GPT-2 tokenizer as a third comparison alongside character- and word-based tokenization.

```{r}
#| label: books-and-tokenizer
baseurl <- "https://www.gutenberg.org/cache/epub/"

books <- data.frame(
  code  = c("84",    "64317",       "11",              "1513",        "76",
            "219",          "2591",         "2148",           "36",              "829"),
  title = c("Frankenstein", "GreatGatsby", "AliceWonderland", "RomeoJuliet", "HuckFinn",
            "HeartDarkness", "GrimmsTales", "EdgarAllenPoe", "WarOfTheWorlds", "GulliversTravels"),
  label = c("Frankenstein", "The Great Gatsby", "Alice in Wonderland", "Romeo and Juliet", 
            "Adventures of Huck Finn", "Heart of Darkness", "Grimm's Tales", "Edgar Allan Poe", 
            "War of the Worlds", "Gullivers Travels")) 
  
books %>%
    flextable()

# Load GPT-2 tokenizer from local file (run gpt2$save("tokenizer.json") once to create it)
gpt2 <- tok::tokenizer$from_file("tokenizer.json")
```

---

# Part 1: Total Characters, Words, and Tokens

For each book, we count total characters (including spaces, punctuation, and newlines), whitespace-delimited words, and GPT-2 tokens.

```{r}
#| label: part1-counts
#| cache: true
results <- vector("list", nrow(books))

for (i in seq_len(nrow(books))) {
  fullurl <- paste0(baseurl, books$code[i], "/pg", books$code[i], ".txt")
  text    <- readLines(fullurl, encoding = "UTF-8") |> paste(collapse = "\n")

  results[[i]] <- data.frame(
    title   = books$title[i],
    n_chars = nchar(text),
    n_words = length(strsplit(trimws(text), "\\s+")[[1]]),
    n_token = length(gpt2$encode(text)$ids)
  )
}

counts <- bind_rows(results)
```

```{r}
#| label: tbl-part1
counts %>%
  flextable() %>%
  flextable::set_caption("Total character, word, and token counts for 10 Project Gutenberg books")
```

Token counts track word counts fairly closely. GPT-2 subword tokenization produces more tokens than words, since punctuation and uncommon words often split into multiple pieces.

---

# Part 2: Unique and Total Counts

Beyond totals, how many *distinct* characters, words, and tokens appear in each book? This is a rough measure of vocabulary richness.

```{r}
#| label: part2-unique
#| cache: true
full_results <- vector("list", nrow(books))

for (i in seq_len(nrow(books))) {
  fullurl <- paste0(baseurl, books$code[i], "/pg", books$code[i], ".txt")
  text    <- readLines(fullurl, encoding = "UTF-8") |> paste(collapse = "\n")

  chars  <- strsplit(text, "")[[1]]
  words  <- strsplit(trimws(text), "\\s+")[[1]]
  tokens <- gpt2$encode(text)$ids

  full_results[[i]] <- data.frame(
    title        = books$title[i],
    total_chars  = length(chars),
    unique_chars = length(unique(chars)),
    total_words  = length(words),
    unique_words = length(unique(words)),
    total_token  = length(tokens),
    unique_token = length(unique(tokens))
  )
}

full_counts <- bind_rows(full_results)
```

```{r}
#| label: tbl-part2
full_counts %>%
  flextable() %>%
  flextable::set_caption("Total and unique counts by tokenization scheme for 10 books")
```

```{r}
#| label: fig-part2
#| fig-cap: "Total (A) and unique (B) counts across tokenization schemes for 10 Project Gutenberg books. Points are individual books; crossbar marks the mean."
#| fig-width: 10
#| fig-height: 3.5

# Reshape to long format for plotting
long_counts <- full_counts |>
  pivot_longer(
    cols      = -title,
    names_to  = c("measure", "scheme"),
    names_sep = "_",
    values_to = "count"
  ) |>
  mutate(
    scheme  = factor(tools::toTitleCase(scheme), levels = c("Chars", "Words", "Token")),
    measure = if_else(measure == "total", "A) Total", "B) Unique")
  )

ggplot(long_counts, aes(x = scheme, y = count)) +
  geom_jitter(width = 0.08, size = 2.5,
              color = "dodgerblue2", alpha = 0.8) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.6,
               linewidth = 0.8, color = "blue") +
  scale_y_continuous(labels = scales::label_scientific()) +
  facet_wrap(~ measure, scales = "free_y") +
  labs(x = NULL, y = "Count") +
  faceted +
  theme(strip.background = element_blank(),
        strip.text        = element_text(face = "bold"))
```

Unique character counts are tightly clustered — most books draw from roughly the same set of ASCII characters. Unique word and token counts spread out more, reflecting differences in author vocabulary and subject matter.

---

# Part 3: Unique by Total Counts

Do longer books use a wider vocabulary? We examine the relationship between total and unique counts across the three tokenization schemes.

```{r}
#| label: fig-part3
#| fig-cap: "Relationship between unique and total counts for each tokenization scheme across 10 books."
#| fig-width: 12
#| fig-height: 3.5

# Reshape: one row per book per scheme, with columns for total and unique
scatter_data <- full_counts |>
  select(title,
         chars_total  = total_chars,  chars_unique  = unique_chars,
         words_total  = total_words,  words_unique  = unique_words,
         token_total  = total_token,  token_unique  = unique_token) |>
  pivot_longer(
    cols      = -title,
    names_to  = c("scheme", "measure"),
    names_sep = "_"
  ) |>
  pivot_wider(names_from = measure, values_from = value) |>
  mutate(scheme = factor(tools::toTitleCase(scheme),
                         levels = c("Chars", "Words", "Token")))

# Per-scheme correlation labels
cor_labels <- scatter_data |>
  group_by(scheme) |>
  summarise(r = cor(unique, total), .groups = "drop") |>
  mutate(label = sprintf("r = %.2f", r))

ggplot(scatter_data, aes(x = unique, y = total)) +
  geom_point(size = 3, color = "dodgerblue2") +
  geom_text(data = cor_labels, aes(label = label),
            x = -Inf, y = Inf, hjust = -0.2, vjust = 1.5,
            fontface = "bold", inherit.aes = FALSE) +
  scale_x_continuous(labels = scales::label_scientific()) +
  scale_y_continuous(labels = scales::label_scientific()) +
  facet_wrap(~ scheme, scales = "free") +
  labs(x = "Unique", y = "Total") +
  faceted +
  theme(strip.background = element_blank(),
        strip.text        = element_text(face = "bold"))
```

Characters show essentially no relationship between unique and total counts — the character set saturates quickly regardless of book length. Word and token unique counts track total counts more closely, suggesting longer books do tend to draw on larger vocabularies.

One book appears as an outlier in unique character count. Let's identify it.

```{r}
#| label: outlier
full_counts |>
  slice_max(unique_chars, n = 1) |>
  select(title, unique_chars) %>%
  flextable()
```

The Edgar Allen Poe collection contains an unusually large number of distinct characters. We can inspect the full character set to understand why.

```{r}
#| label: outlier-chars
text_poe <- readLines(
  "https://www.gutenberg.org/cache/epub/2148/pg2148.txt",
  encoding = "UTF-8"
) |> paste(collapse = "\n")

sort(unique(strsplit(text_poe, "")[[1]]))
```

The Poe collection includes accented and non-Latin characters — likely from French or other language quotations embedded in the text — which inflates the unique character count relative to the other books.

# Addendum

Can we query the corpus with a sentence, passage, or keywords, and get back a list of similarity to the books in the corpus? We will embed each book as the average of its chunk embeddings, embed the query, then rank books by cosine similarity to the query.

```{r}
#| label: semantic-search
library(reticulate)

# helper: split text into ~500-word chunks
chunk_text <- function(text, size = 500) {
  words  <- strsplit(text, "\\s+")[[1]]
  starts <- seq(1, length(words), by = size)
  purrr::map_chr(starts, \(s) paste(words[s:min(s + size - 1, length(words))], collapse = " "))
}

# load model (downloads ~90MB on first run)
st    <- import("sentence_transformers")
model <- st$SentenceTransformer("all-MiniLM-L6-v2")

# fetch all book texts
book_texts <- purrr::map(books$code, \(code) {
  paste0(baseurl, code, "/pg", code, ".txt") |>
    readLines(encoding = "UTF-8") |>
    paste(collapse = "\n")
})

# embed each book as average of chunk embeddings (save to disk to avoid recomputing)
emb_path <- "book_emb.rds"

if (file.exists(emb_path)) {
  book_emb <- readRDS(emb_path)
} else {
  book_emb <- purrr::map(book_texts, \(text) {
    chunks <- chunk_text(text)
    embs   <- model$encode(chunks)
    colMeans(embs)
  })
  saveRDS(book_emb, emb_path)
}

# embed query and compute cosine similarity to each book
query     <- "revenge and obsession"
query_emb <- model$encode(query)

cos_sim <- purrr::map_dbl(book_emb, \(emb) {
  sum(query_emb * emb) / (sqrt(sum(query_emb^2)) * sqrt(sum(emb^2)))
})

out <- tibble(title = books$label, cos_sim = cos_sim) |>
  arrange(desc(cos_sim)) 

out %>%
    flextable()
```

```{r}
#| fig-height: 4
#| fig-width: 7
ggplot(out, aes(cos_sim, reorder(title, cos_sim))) +
    geom_label(aes(label=round(cos_sim,2)),
               fill="dodgerblue2",
               color="white",
               size=4) +
    labs(x="Cosine similarity",
         y="",
         title=glue::glue("Novel similarity to: {query}"))
```
Now let's wrap the query into a function for easier interactive work. 

```{r}
semantic_search <- function(query) {
  q_emb <- model$encode(query)
  cos_sim <- purrr::map_dbl(book_emb, \(emb) {
    sum(q_emb * emb) / (sqrt(sum(q_emb^2)) * sqrt(sum(emb^2)))
  })
  out <- tibble(title = books$label, cos_sim = cos_sim) |> arrange(desc(cos_sim))
  out
  ggplot(out, aes(cos_sim, reorder(title, cos_sim))) +
    geom_label(aes(label=sub("^0", "", as.character(round(cos_sim, 2)))),
               fill="navy",
               color="white",
               #color="dodgerblue",
               fontface="bold",
               size=6,
               label.padding=unit(.1, "lines")) +
      scale_x_continuous(sec.axis=dup_axis()) +
    labs(x="Cosine similarity",
         y="",
         title=glue::glue("Novel similarity to:\n{query}")) +
      theme(axis.title.x.top=element_blank())
}
```

```{r}
semantic_search("childhood wonder and adventure")
```
I would have expected Alice, then Huck Finn, then Gullivers. I can see the query relating to a deeper theme in Gatsby. 


```{r}
semantic_search("Love and death")
```
Shouldn't the first two be Romeo and Gatsby? I can see a deeper tie-in to Frankenstein. 

```{r}
semantic_search("space the final frontier")
```
Looks good

```{r}
semantic_search("Oh how I would love to do the impossible!")
```
I would expect higher rankings for Grimm, Gatsby, and Frankenstein

```{r}
semantic_search("Why is it so easy for people to hate each other?")
```
This query was to see if it picked up on the family feud in Huck Finn. I can see the other rankings but not sure why Alice is up there. 





















