---
title: "Project 1: Three Tokenization Schemes"
subtitle: "50 ML Projects to Understand LLMs — Chapter 2"
author: "R translation of Mike X Cohen's original Python notebook"
date: today
format:
  html:
    toc: true
    code-fold: false
    theme: cosmo
execute:
  warning: false
  message: false
---

*Using the code without reading the book may lead to confusion or errors.*

---

```{r global_options, include=F, warning=F, message=F, echo=F, error=F}
library(here)
library(knitr)
source("../../../prep (May 2025).r")
```

---

# Part 1: Character-Based Tokenization

```{r}
#| label: char-setup
# Start with a text string
txt <- "The way you do anything is the way you do everything."

# Split into individual characters
characters <- strsplit(txt, "")[[1]]

characters

```

```{r}

cat("The full text:\n", txt, "\n\n")
cat("As a vector of characters:\n", characters)
print(characters)

# Build the vocabulary (sorted unique characters)
char_vocab <- sort(unique(characters))

cat(sprintf(
  "\nThere are %d characters, %d of which are unique.\n",
  length(characters), length(char_vocab)
))

cat("\nThe vocabulary is:\n", char_vocab)

```
Create our own token index

```{r}
#| label: char-vocab-table
# Display token–index mapping in a table
char_vocab_tbl <- data.frame(Index = seq_along(char_vocab),
                             Token = char_vocab) |> 
    flextable()

char_vocab_tbl

```

Map each character to its index in the vocab

```{r}
#| label: fig-char-scatter
#| fig-cap: "Character-level token indices for the example sentence."
#| fig-width: 12
#| fig-height: 4
#| 
df_char <- data.frame(
  position = seq_along(tokens_char),
  character = characters,
  token    = match(characters, char_vocab))

df_char %>%
    flextable()

```

```{r}

ggplot(df_char, aes(x = position, y = token)) +
  geom_point(shape = 22, size = 4, fill = "#b3b3e5", color = "black") +
  scale_y_continuous(
    breaks   = seq_along(char_vocab) - 1L,
    sec.axis = sec_axis(~ ., breaks = seq_along(char_vocab) - 1L,
                        labels = char_vocab)
  ) +
  labs(x = "Character index", y = NULL) +
  theme_classic() +
  theme(panel.grid.major.y = element_line(linetype = "dashed", color = "grey70"))
```

Not sure what the point is of this graph. 

```{r}

char_count <- df_char |>
  count(character) |>
  mutate(index = match(character, char_vocab)) |>
  arrange(index)

char_count

```

```{r}

ggplot(char_count, aes(reorder(character, n), n)) +
    geom_col(fill="dodgerblue2",
             color="blue",
             alpha=.8,
             width=.3) +
    scale_y_continuous(breaks=1:10,
                       sec.axis=dup_axis()) +
    labs(x="",
         y="",
         title="Count of unique characters") +
    theme(axis.title.y=element_text(angle=0, vjust=.55))

```


---

# Part 2: Word-Based Tokenization

```{r}
#| label: word-setup
# Split the text into words on whitespace
words <- strsplit(txt, "\\s+")[[1]]

cat("The list of words is:\n")
print(words)

# Build the word vocabulary
word_vocab <- sort(unique(words))

cat(sprintf(
  "\nThere are %d words, %d of which are unique.\n",
  length(words), length(word_vocab)
))

cat("\nThe vocab is:\n")
print(word_vocab)
```

```{r}
#| label: word-encode
# Encode: map each word to its vocab index (0-indexed)
tokens_word <- match(words, word_vocab) - 1L
print(tokens_word)
```

```{r}
#| label: fig-word-scatter
#| fig-cap: "Word-level token indices for the example sentence."
#| fig-width: 10
#| fig-height: 3
df_word <- data.frame(
  position = seq_along(tokens_word) - 1L,
  token    = tokens_word
)

ggplot(df_word, aes(x = position, y = token)) +
  geom_point(shape = 22, size = 5, fill = "#b3b3e5", color = "black") +
  scale_y_continuous(
    breaks   = seq_along(word_vocab) - 1L,
    sec.axis = sec_axis(~ ., breaks = seq_along(word_vocab) - 1L,
                        labels = word_vocab)
  ) +
  labs(x = "Word index", y = NULL) +
  theme_classic() +
  theme(panel.grid.major.y = element_line(linetype = "dashed", color = "grey70"))
```

---

# Part 3: OpenAI's GPT-2 Tokenizer

This section uses the [`tok`](https://cran.r-project.org/package=tok) package,
which provides R bindings to HuggingFace's `tokenizers` library.

```{r}
#| label: gpt2-load
library(tok)

# Use tok::tokenizer to reference the class, not any instance variable
tokenizer <- tok::tokenizer$from_pretrained("gpt2")
tokenizer$get_vocab_size()
```

```{r}
#| label: gpt2-encode
gpt2_tokens <- tokenizer$encode(txt)$ids

# Print each token alongside its decoded string
for (token in gpt2_tokens) {
  decoded <- tokenizer$decode(token)
  cat(sprintf('Token %4d is "%s"\n', token, decoded))
}
```

```{r}
#| label: gpt2-leading-spaces
# Demonstrate how leading spaces affect tokenization
word1 <- " Mike"
word2 <- "Mike"

for (w in c(word1, word2)) {
  ids <- tokenizer$encode(w)$ids
  cat(sprintf('Token for "%s" is %s\n', w, paste(ids, collapse = ", ")))
}
```

```{r}
#| label: gpt2-counts
cat(sprintf(
  "There are %d tokens, %d of which are unique.\n",
  length(gpt2_tokens), length(unique(gpt2_tokens))
))
```

```{r}
#| label: fig-comparison
#| fig-cap: "Comparison of total and unique token counts across the three tokenization schemes."
#| fig-width: 10
#| fig-height: 3.5
counts <- data.frame(
  scheme = factor(
    rep(c("Characters", "Words", "GPT-2"), 2),
    levels = c("Characters", "Words", "GPT-2")
  ),
  type  = rep(c("Total", "Unique"), each = 3),
  count = c(
    length(characters), length(words), length(gpt2_tokens),
    length(char_vocab), length(word_vocab), length(unique(gpt2_tokens))
  )
)

ggplot(counts, aes(x = scheme, y = count)) +
  geom_col() +
  facet_wrap(~ type) +
  labs(x = NULL, y = "Count") +
  theme_classic()
```

# Addendum

## Line-based tokenization

```{r}
dig <- "Between my finger and my thumb
The squat pen rests; snug as a gun.
 
Under my window, a clean rasping sound
When the spade sinks into gravelly ground:
My father, digging. I look down
 
Till his straining rump among the flowerbeds
Bends low, comes up twenty years away
Stooping in rhythm through potato drills
Where he was digging.
 
The coarse boot nestled on the lug, the shaft
Against the inside knee was levered firmly.
He rooted out tall tops, buried the bright edge deep
To scatter new potatoes that we picked,
Loving their cool hardness in our hands.
 
By God, the old man could handle a spade.
Just like his old man.
 
My grandfather cut more turf in a day
Than any other man on Toner’s bog.
Once I carried him milk in a bottle
Corked sloppily with paper. He straightened up
To drink it, then fell to right away
Nicking and slicing neatly, heaving sods
Over his shoulder, going down and down
For the good turf. Digging.
 
The cold smell of potato mould, the squelch and slap
Of soggy peat, the curt cuts of an edge
Through living roots awaken in my head.
But I’ve no spade to follow men like them.
 
Between my finger and my thumb
The squat pen rests.
I’ll dig with it." 

```

```{r}

dig_df <- dig %>%
    tibble(text = str_split(dig, "\n")[[1]]) |>
  mutate(
    is_blank = str_trim(text) == "",
    stanza = cumsum(is_blank & lag(is_blank | row_number() == 1, default = TRUE)) + 1) |>
  filter(!is_blank) |>
  mutate(line = row_number()) |>
  select(stanza, line, text)

dig_df 

```


