---
title: "Three Tokenization Schemes"
subtitle: "50 ML Projects to Understand LLMs — Chapter 2, Project 1"
author: "Dan Killian, with Posit AI<br>an R translation of Mike X Cohen's original Python notebook"
date: today
format:
  html:
    toc: true
    code-fold: false
    theme: cosmo
execute:
  warning: false
  message: false
---

*Using the code without reading the book may lead to confusion or errors.*

---

```{r global_options, include=F, warning=F, message=F, echo=F, error=F}
library(here)
library(knitr)
source("../../../prep (May 2025).r")
```

---

# Part 1: Character-Based Tokenization

```{r}
#| label: char-setup

# Start with a text string
txt <- "The way you do anything is the way you do everything."

# Split into individual characters
characters <- strsplit(txt, "")[[1]]

# Build the vocabulary (sorted unique characters)
char_vocab <- sort(unique(characters))
```

There are `r length(characters)` characters, `r length(char_vocab)` of which are unique

The full text:

`r txt`

As a vector of characters: 

`r characters`

The vocabulary:  
`r char_vocab`


Create our own character token index

```{r}
#| label: char-vocab-table
# Display token–index mapping in a table
char_vocab_tbl <- data.frame(Index = seq_along(char_vocab),
                             Token = char_vocab) |> 
    flextable()

char_vocab_tbl

```

Map each character to its index in the vocab

```{r}
#| label: fig-char-scatter
#| fig-cap: "Character-level token indices for the example sentence."
#| fig-width: 12
#| fig-height: 4
#| 
df_char <- data.frame(
  position = seq_along(characters),
  character = characters,
  token    = match(characters, char_vocab))

df_char %>%
    flextable()

```

How often do characters appear?

```{r}

char_count <- df_char |>
  count(character) |>
  mutate(index = match(character, char_vocab)) |>
  arrange(index)

char_count %>%
    flextable()

```

```{r}

theme_set(base_ppt)

ggplot(char_count, aes(reorder(character, -n), n)) +
    geom_col(fill="dodgerblue2",
             color="blue",
             alpha=.8,
             width=.3) +
    scale_y_continuous(breaks=1:10,
                       sec.axis=dup_axis()) +
    labs(x="",
         y="",
         title="Count of unique characters") +
    theme(axis.title.y=element_text(angle=0, vjust=.55))

```

---

# Part 2: Word-Based Tokenization

```{r}
#| label: word-setup

# Split the text into words on whitespace
words <- strsplit(txt, "\\s+")[[1]] # \\s+ means one or more whitespace characters

# Build the word vocabulary
word_vocab <- sort(unique(words))

```

There are `r length(words)` words, `r length(word_vocab)` of which are unique

The text:  
`r words`

The vocab:  
`r word_vocab`

Note the duplication of 'the' based on capitalization. The period after 'everything' may also mess things up. 


Create our own word token index

```{r}
#| label: word-vocab-table
# Display token–index mapping in a table
wrd_vocab_tbl <- data.frame(Index = seq_along(word_vocab),
                             Token = word_vocab) |> 
    flextable()

wrd_vocab_tbl

```

Map our token index back to the corpus

```{r}

df_wrd <- tibble(position = seq_along(words),
                 word=words,
                 token    = match(words, word_vocab))

df_wrd %>%
    flextable()


```
How often does each word show up?

```{r}
wrd_count <- df_wrd |>
  count(word) |>
  mutate(index = match(word, word_vocab)) |>
  arrange(desc(n))

wrd_count %>%
    flextable()
```

```{r}

ggplot(wrd_count, aes(reorder(word, -n), n)) +
    geom_col(fill="dodgerblue2",
             color="blue",
             alpha=.8,
             width=.3) +
    scale_y_continuous(breaks=1:10,
                       sec.axis=dup_axis()) +
    labs(x="",
         y="",
         title="Count of unique words") +
    theme(axis.title.y=element_text(angle=0, vjust=.55))

```

---

# Part 3: OpenAI's GPT-2 Tokenizer

This section uses the [`tok`](https://cran.r-project.org/package=tok) package,
which provides R bindings to HuggingFace's `tokenizers` library.

```{r}
#| label: gpt2-load
library(tok)

# Use tok::tokenizer to reference the class, not any instance variable
tokenizer <- tok::tokenizer$from_pretrained("gpt2")
tokenizer$get_vocab_size()
```
GPT2 has an existing corpus comprising fifty thousand tokens. 

```{r}
#| label: gpt2-encode
gpt2_tokens <- tokenizer$encode(txt)$ids

# saveRDS(gpt2_tokens, "gpt2_tokens.rds")

# Print each token alongside its decoded string
for (token in gpt2_tokens) {
  decoded <- tokenizer$decode(token)
  cat(sprintf('Token %4d is "%s"\n', token, decoded))
}
```

```{r}
#| label: gpt2-leading-spaces
# Demonstrate how leading spaces affect tokenization
word1 <- " Mike"
word2 <- "Mike"

for (w in c(word1, word2)) {
  ids <- tokenizer$encode(w)$ids
  cat(sprintf('Token for "%s" is %s\n', w, paste(ids, collapse = ", ")))
}
```
Why allow a separate token for a white space before a name? Is that an oversight?

```{r}
#| label: gpt2-counts
cat(sprintf(
  "There are %d tokens, %d of which are unique.\n",
  length(gpt2_tokens), length(unique(gpt2_tokens))
))
```

```{r}
#| label: fig-comparison
#| fig-cap: "Comparison of total and unique token counts across the three tokenization schemes."
#| fig-width: 10
#| fig-height: 3.5
counts <- data.frame(
  scheme = factor(
    rep(c("Characters", "Words", "GPT-2"), 2),
    levels = c("Characters", "Words", "GPT-2")
  ),
  type  = rep(c("Total", "Unique"), each = 3),
  count = c(
    length(characters), length(words), length(gpt2_tokens),
    length(char_vocab), length(word_vocab), length(unique(gpt2_tokens))
  )
)

ggplot(counts, aes(x = scheme, y = count)) +
  geom_col(width=.4,
           fill="dodgerblue2",
           color="blue",
           alpha=.6) +
  facet_wrap(~ type) +
  labs(x = NULL, y = "Count") +
  faceted
```

# Addendum: New corpus

```{r}
dig <- "Between my finger and my thumb
The squat pen rests; snug as a gun.
 
Under my window, a clean rasping sound
When the spade sinks into gravelly ground:
My father, digging. I look down
 
Till his straining rump among the flowerbeds
Bends low, comes up twenty years away
Stooping in rhythm through potato drills
Where he was digging.
 
The coarse boot nestled on the lug, the shaft
Against the inside knee was levered firmly.
He rooted out tall tops, buried the bright edge deep
To scatter new potatoes that we picked,
Loving their cool hardness in our hands.
 
By God, the old man could handle a spade.
Just like his old man.
 
My grandfather cut more turf in a day
Than any other man on Toner’s bog.
Once I carried him milk in a bottle
Corked sloppily with paper. He straightened up
To drink it, then fell to right away
Nicking and slicing neatly, heaving sods
Over his shoulder, going down and down
For the good turf. Digging.
 
The cold smell of potato mould, the squelch and slap
Of soggy peat, the curt cuts of an edge
Through living roots awaken in my head.
But I’ve no spade to follow men like them.
 
Between my finger and my thumb
The squat pen rests.
I’ll dig with it." 

```

## Word-based tokenization

```{r}
#| label: dig-setup
# Split the text into words on whitespace
dig_words <- strsplit(dig, "\\s+") [[1]] # \\s+ means one or more whitespace characters

# Build the word vocabulary
dig_word_vocab <- sort(unique(dig_words))

```

There are `r length(dig_words)` words, `r length(dig_word_vocab)` of which are unique

The vocabulary:  
`r dig_word_vocab`


Create our own word token index

```{r}
#| label: dig-vocab-table
# Display token–index mapping in a table
dig_vocab_tbl <- data.frame(Index = seq_along(dig_word_vocab),
                             Token = dig_word_vocab) |> 
    flextable()

dig_vocab_tbl

```


Map our token index back to the corpus

```{r}
dig_df <- tibble(position = seq_along(dig_words),
                 word=dig_words,
                 token    = match(dig_words, dig_word_vocab))

dig_df %>%
    flextable()
```

Count of unique words

```{r}
dig_count <- dig_df |>
  count(word) |>
  mutate(index = match(word, dig_word_vocab)) |>
  arrange(desc(n))

dig_count %>%
    flextable()
```

```{r}
#| label: digging word count
#| fig-cap: "Word count: Digging, Seamus Heaney"
#| fig-width: 4
#| fig-height: 18
ggplot(dig_count, aes(n, reorder(word, n))) +
    geom_col(fill="dodgerblue2",
             color="blue",
             alpha=.8,
             width=.3) +
    scale_x_continuous(breaks=1:10,
                       sec.axis=dup_axis()) +
    labs(x="",
         y="",
         title="Count of unique words") +
    theme(axis.title.y=element_text(angle=0, vjust=.55))
```

## Line-based tokenization

```{r}
#| label: dig-setup-line
# Split the text into words on whitespace
dig_lines <- strsplit(dig, "\n") [[1]] |>
    str_subset("\\S")

dig_stanzas <- strsplit(dig, "\n \n")[[1]]   

```

There are `r length(dig_lines)` lines across `r length(dig_stanzas)` stanzas

```{r}

dig_df <- tibble(text = str_split(dig, "\n")[[1]]) |>
    mutate(is_blank = str_trim(text) == "",
           stanza = cumsum(lag(is_blank, default = FALSE)) + 1) |>
    filter(!is_blank) |>
    mutate(line = row_number()) |>
    select(stanza, line, text)

dig_df |>
    flextable()

```
How might we develop descriptive statistics of lines and stanzas?

```{r}
line_df <- dig_df |>
  mutate(
    word_count = str_count(text, "\\S+"),
    char_count = nchar(text)
  )

line_df %>%
    flextable()
```
```{r}
#| label: tbl-lines
#| tbl-cap: "Characteristics of poem lines"

line_desc <- describe(line_df[,4:5]) %>%
    mutate(token_type = c("Words","Characters")) %>%
    .[,c(14,3, 5,4, 13, 8:9)] 


line_desc %>%
    flextable() %>%
    flextable::set_caption("Characteristics of poem lines")
```

```{r}
stanzas_df <- tibble(stanza=1:8,
                     text=dig_stanzas) |>
    group_by(stanza) %>%
    mutate(word_count = str_count(text, "\\S+"),
           char_count = nchar(text))

stanzas_df %>%
    flextable()
```

```{r}
#| label: tbl-stanzas
#| tbl-cap: "Characteristics of poem stanzas"

stanz_desc <- describe(stanzas_df[,3:4]) %>%
    mutate(token_type = c("Words","Characters")) %>%
    .[,c(14,3, 5,4, 13, 8:9)] 


stanz_desc %>%
    flextable() %>%
    flextable::set_caption("Characteristics of poem stanzas")
```



## Embeddings

``` {r}

# RUN ONCE, THEN LOAD THE EMBEDDINGS

# library(openai)

# Reads OPENAI_API_KEY from your environment automatically
# embeddings <- create_embedding(
#   model = "text-embedding-3-small",
#   input = dig_df$text
# )
# 
# d <- embeddings$data
# 
# # Extract the matrix of vectors
# # embed_matrix <- sapply(embeddings$data, \(x) x$embedding)
#     
# embed_matrix <- do.call(rbind, d$embedding)
# 
# # Result: a matrix where each column is the embedding for one line
# 
# saveRDS(embed_matrix, "embeddings.rds")

embed_matrix <- readRDS("embeddings.rds")

```

### Cosine similarities between lines

```{r}
library(lsa)

sim_matrix <- cosine(t(embed_matrix))
```

```{r}
sim_matrix |>
  as.data.frame() |>
  setNames(seq_len(ncol(sim_matrix))) |>
  mutate(line_i = row_number()) |>
  tidyr::pivot_longer(-line_i, names_to = "line_j", values_to = "similarity") |>
  mutate(line_j = as.integer(line_j)) |>
  ggplot(aes(line_i, line_j, fill = similarity)) +
  geom_tile() +
  scale_fill_distiller(palette = "Blues", direction = 1) +
  labs(x = "Line", y = "Line", fill = "Cosine\nsimilarity")
```
A couple lines overlap and have similarity of 1 or close to it


```{r}
sim_df <- which(upper.tri(sim_matrix), arr.ind = TRUE) |>
  as.data.frame() |>
  setNames(c("i", "j")) |>
  mutate(
    line_i     = dig_df$text[i],
    line_j     = dig_df$text[j],
    similarity = sim_matrix[cbind(i, j)]
  )

sim_df <- sim_df %>%
#    group_by(i) %>%
    arrange(i, desc(similarity))
```


```{r}
# Extract upper triangle values
sim_vals <- sim_matrix[upper.tri(sim_matrix)]

sim_desc <- psych::describe(sim_vals) %>%
    .[,c(1:3, 5,4, 13, 8:9)] 

sim_desc %>%
    flextable
```


```{r}
sim_df <- which(upper.tri(sim_matrix), arr.ind = TRUE) |>
  as.data.frame() |>
  setNames(c("i", "j")) |>
  mutate(
    line_i     = dig_lines[i],
    line_j     = dig_lines[j],
    similarity = sim_matrix[cbind(i, j)]
  ) %>%
    select(1,3,2,4,5) %>%
    arrange(i)

sim_df |> flextable()
```

### Stanza cohesion

Are lines within stanzas different lines in different stanzas?

```{r}
sim_df |>
  mutate(
    stanza_i = dig_df$stanza[i],
    stanza_j = dig_df$stanza[j],
    same_stanza = stanza_i == stanza_j
  ) |>
  group_by(same_stanza) |>
  summarise(mean_sim = mean(similarity)) %>%
    flextable()
```

No not really 

### Semantic search

How might we query a poem? 

```{r}
#| cache: true

library(openai)

# Embed the query
query <- "manual labour and memory"

query_vec <- create_embedding(model = "text-embedding-3-small",
                              input = query)$data$embedding[[1]] |> 
    unlist()

# Cosine similarity between query and each line
cos_sim <- function(a, b) sum(a * b) / (sqrt(sum(a^2)) * sqrt(sum(b^2)))

out <- tibble(
  stanza = dig_df$stanza,
  line = dig_df$line,
  text = dig_df$text,
  similarity = apply(embed_matrix, 1, \(row) cos_sim(query_vec, row))
) |>
  arrange(desc(similarity))

out %>%
    flextable() %>%
    flextable::set_caption("query: manual labor and memory")
```

Write a function to allow us to query the poem with text, and get back the poem lines with highest similarity

```{r}
search_poem <- function(query, n = 5, lines = dig_df$text, matrix = embed_matrix) {
  cos_sim <- function(a, b) sum(a * b) / (sqrt(sum(a^2)) * sqrt(sum(b^2)))
  
  query_vec <- create_embedding(
    model = "text-embedding-3-small",
    input = query
  )$data$embedding[[1]] |> unlist()
  
  tibble(
    line       = lines,
    stanza     = dig_df$stanza,
    similarity = apply(matrix, 1, \(row) cos_sim(query_vec, row))
  ) |>
    slice_max(similarity, n = n)              # <- returns tibble, not flextable
}
```


```{r}
# q1 <- search_poem("digging and earth")

# saveRDS(q1, "q1.rds")

q1 <- readRDS("q1.rds")

q1 %>% 
    flextable()

```

```{r}
# q2 <- search_poem("Japan")

# saveRDS(q2, "q2.rds")

q2 <- readRDS("q2.rds")

q2 %>% 
    flextable()

```

```{r}
# q3 <- search_poem("ennui")

# saveRDS(q3, "q3.rds")

q3 <- readRDS("q3.rds")

q3 %>% 
    flextable()
```


```{r}
# q4 <- search_poem("reputation")

# saveRDS(q1, "q4.rds")

q4 <- readRDS("q4.rds")

q4 %>% 
    flextable()

```


```{r}
# q5 <- search_poem("memory")

# saveRDS(q5, "q5.rds")

q5 <- readRDS("q5.rds")

q5 %>% 
    flextable()

```


```{r}
# q6 <- search_poem("jealousy")

# saveRDS(q6, "q6.rds")

q6 <- readRDS("q6.rds")

q6 %>% 
    flextable()

```

Interesting








